Handling missing data in smartphone location logs

### 2017-11-25 thesisReport.Rmd
problem description comments

In the example data set the distribution of $a$ is highly right skewed, with a median of 28, $\mu = 127$ and the maximum value at 26 km. @palmius_detecting_2017 note that in their Android based custom logs inaccurate location values are interspersed between more accurate location values at higher sample rates per hour. We observe similar patterns in secondary logs. Figure 1 shows how accuracy tends to vary as a function of user behaviour, time and location. There are several recurring low-accuracy points, most likely cell-phone triangulation towers.

@@

Over 54% of the data is missing for the entire duration of the log. However,this is misleading as there are several long periods with no measurements whatsoever. The structure of missingness of a day with measurements is shown below.

```{r, measurementsPerDay,echo = FALSE,fig.cap="Example of missing data over a day. The x-axis denotes time, the y-axis shows how many measurements are made and each point is a five minute window. For this day there were several periods with no information. These points are filled with red and lie on the x-axis.",out.width = '100%'}
#fix width
include_graphics("img/missingBoaz5minExample.png")

@@

## Current methods for imputing & filtering spatiotemporal data

As we have mentioned before, @jankowska_framework_2015 point out that there is little transparency regarding decisions of how to deal with missing data in GPS mobility logs. We suspect this is because the highly interdisciplinary nature of the problem means researchers are unaware of potential solutions. For this reason, we consider it important to briefly discuss a few methods one could consider in order to deal with measurement inaccuracy and missing data problems in spatiotemporal data. In doing so, we will argue that state space models (SSMs) and spatiotemporal imputation methods, which are extensively used to model missing and noisy spatiotemporal data, are not well suited to deal with messy human mobility patterns. Moreover, we discuss in detail two approaches by @palmius_detecting_2017 and @barnett_inferring_2016 which deal explicitly with missing data in mobility patterns from smartphone GPS logs.

### General Spatiotemporal Methods

@@

Four imputation methods were selected in order to cover techniques applied in the literature. Briefly, the mean imputation method described by @palmius_detecting_2017, the model developed by @barnett_inferring_2016 as well as an adapted FarOut model [@sadilek_far_2016]. Simple linear interpolation was used as a benchmark model.

@@

The evaluation on computed measures involved calculating a mobility trace following the rectangual method of @rhee_human_2007 for each imputed dataset. Like @barnett_inferring_2016 we calculate bias by substracting the estimated measure under each approach for the same measure calculated on the full data. For simulation-based imputation approaches a mean value over 100 samples was taken.

## Results

## Discussion

>>>>

In the example data set the distribution of $a$ is highly right skewed, with a median of 28, $\mu = 127$ and the maximum value at 26 km. @palmius_detecting_2017 note that in their Android based custom logs inaccurate location values are interspersed between more accurate location values at higher sample rates per hour. We observe similar patterns in secondary logs. Figure 1 shows how accuracy can vary as a function of user behaviour, time and location. Most notably, accuracy tends to be lowest during movement. Stationary accuracy varies depending on phone battery level, wifi connection and user phone use. There are several recurring low-accuracy points, most likely cell-phone triangulation towers.

@@

Over 54% of the data is missing for the entire duration of the log (as shown in Figure X a). However,this is misleading as there are several long periods with no measurements whatsoever. The structure of missingness of a day with measurements is shown in Figure X b. As you can see, there are several long periods over the course of the log for which there are no measurements. Moreover, even during a single day there are continuous periods where there is missing data, mostly during the late hours of the night in this case.

```{r, longMeasurementsPerDay,echo = FALSE,fig.cap="Example of missing data over the entire duration of the log. The x-axis denotes time, the y-axis shows how many measurements are made and each point is a five minute window. For this day there were several periods with no information. These points are filled with red and lie on the x-axis.",out.width = '100%'}

include_graphics("img/missingdayBoaz5min.png")

@@

```{r, measurementsPerDay,echo = FALSE,fig.cap="Example of missing data over a day. The x-axis denotes time, the y-axis shows how many measurements are made and each point is a five minute window. For this day there were several periods with no information. These points are filled with red and lie on the x-axis.",out.width = '100%'}
#fix width
include_graphics("img/missingBoaz5minExample.png")
```

## Current methods for imputing missing spatiotemporal data

As we have mentioned before, @jankowska_framework_2015 point out that there is little transparency regarding decisions of how to deal with missing data in GPS mobility logs. We suspect this is because the highly interdisciplinary nature of the problem means researchers are unaware of potential solutions. For this reason, we consider it important to briefly discuss a few methods one could consider in order to deal with measurement inaccuracy and missing data problems in spatiotemporal data. In doing so, we will argue that state space models (SSMs) and spatiotemporal imputation methods, which are extensively used to model missing and noisy spatiotemporal data, are not well suited to deal with human mobility patterns. Moreover, we discuss in detail two approaches by @palmius_detecting_2017 and @barnett_inferring_2016 which deal explicitly with missing data in mobility patterns from smartphone GPS logs.

@@

Three imputation methods were selected in order to cover a wide range of techniques applied in the literature:

1. Mean imputation as described by @palmius_detecting_2017.
2. The model developed by @barnett_inferring_2016 using both spatial and temporal proximity.
3. Simple linear interpolation was used as a benchmark model.

@@

The evaluation on computed measures involved calculating a mobility trace following the rectangular method of @rhee_human_2007 for each imputed dataset. Like @barnett_inferring_2016 we calculate bias by substracting the estimated measure under each approach for the same measure calculated on the full data. For simulation-based imputation approaches a mean value over 100 samples was taken.

Each imputation method used each of the three filtering methods as an input. Thus in the end we end up with eight methods to evaluate, three for each filtering method as well as four for each imputation method.
